Supplemental data to ``\emph{k}-mer-based GWAS enhances causal variant and candidate gene discovery in soybean''

### 2023-02-20 additional_files/main_text.tex
Integrate suggestions by JA Sibbesen into the main Results section

We analyzed a total of 13 traits using four different GWAS approaches, one
based on SNP and indel genotypes (Platypus), two based on SV genotypes (Paragraph and
vg), and another based on the presence/absence of \textit{k}-mers.
Because two or three different GWAS analyses were conducted on some traits, a
total of 22 GWAS analyses were performed using each of the four approaches.
Given the large volume of data that this represents, we will focus here on the
most noteworthy results. Readers interested in seeing detailed results on a
per-trait basis are referred to
Supplemental data. In particular, table \ref{signals_table}
provides a summary of the \textit{p}-values obtained by each approach at each
of the loci considered in this study.

@@

We were first interested in assessing the performance of various GWAS
approaches by analyzing the results at loci for which the genes have
already been cloned. These loci provide interesting test cases because
the expected results (causal genes or variants) are known. Table \ref{loci-table} shows
a summary of the results discussed below.

@@

\cite{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Figures
\ref{flower_color_manhattan}, \ref{flower_color_W1_signal}) and the most
significant variants/\textit{k}-mers of all approaches except Platypus identified
the known causal variant (Figures \ref{flower-color-main-figure} and
\ref{flower_color_W1_gene}).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, variants located in the gene proved to be among the most closely
associated, but the causal variant could not be captured given the large size
of this variant.

We should note that when we first analyzed the results of the GWAS on flower
color, we observed that two accessions bearing the haplotype typically
associated with purple flowers were noted as having white flowers. Upon further
analysis, we found that the most likely cause of this discrepancy was a
mismatch between the genotype data and the identity of the accession for which
the phenotypic data was obtained. Indeed, when we compared SNP calls made
directly from the WGS data to those made using the SoySNP50K array, we observed
that these two accessions had concordance rates of only 62\% and 70\%. To
prevent such mismatches from affecting our GWAS analyses, we computed the
concordance between genotype calls made from WGS data and the SoySNP50K array
for the whole population (see the Methods section) and
excluded 24 accessions for which the concordance rate was below 90\%.

@@

% Pubescence color - T locus
\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as
a putative causal variant at this locus. We found signals overlapping this gene
using all four approaches (Figure \ref{pubescence_color_all_manhattan}). In
particular, the regions defined by the top 1\% most associated variants identified by the
Platypus and \textit{k}-mer approaches included Glyma.06g202300 (Figure
\ref{pubescence_color_all_T_signal}). While the most significant
\textit{k}-mers did occur within the sequence of this gene, they did not
correspond to the causal variant (Figure \ref{pubescence_color_all_T_gene}).
Still, the fourth and fifth most strongly associated \textit{k}-mers
corresponded to the causal variant (Figures \ref{pubescence_color_all_T_gene},
\ref{pubescence_color_all_T_kmers}). As for Platypus, the most significant
variant within the sequence of Glyma.06g202300 was the causal
variant, however it was not the most significantly associated variant overall
(Figures \ref{pubescence_color_all_T_signal},
\ref{pubescence_color_all_T_gene}). Neither vg nor Paragraph detected the causal
variant at this locus given that these approaches did not target small indels.

@@

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny and light
tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we
detected signals in this region using all four approaches, but only Platypus
and the \textit{k}-mers identified signals that overlapped Glyma.03g258700
(Figures \ref{pubescence-color-main-figure}a-b,
\ref{pubescence_color_nogray_manhattan}, \ref{pubescence_color_nogray_Td_signal}).
While the most significant \textit{k}-mer identified the causal variant documented
at this locus \citep{yan2020}, the fourth most significantly associated variant
found by Platypus corresponded to this variant (Figures
\ref{pubescence-color-main-figure}b-d, \ref{pubescence_color_nogray_Td_gene}).
Interestingly, some of the accessions bearing the haplotype associated with tawny pubescence
displayed light tawny pubescence, suggesting that there may be more than one causal
variant at this locus (Figure \ref{pubescence-color-main-figure}e).

@@

% Seed coat color - I locus
The causal variant at the \textit{I} locus for seed coat color is a complex SV
that takes the form of a 10.91-kb inverted duplication of three chalcone
synthase (CHS) genes \citep[CHS1, CHS3 and CHS4;][]{tuteja2008}. The presence of
this inverted duplication results in the silencing of all CHS genes in the seed
coat specifically and thus in the absence of seed coat pigmentation
\citep{tuteja2009}.  We did identify signals overlapping this causal variant
using all four GWAS approaches (Figures \ref{seed_coat_color_all_manhattan},
\ref{seed_coat_color_all_I_signal}), however none of the approaches identified
the causal variant. This failure to detect the causal variant is likely due
to its complexity, which makes it difficult to detect and genotype. While it
is likely that at least some of the significantly associated \textit{k}-mers
actually are derived from this causal variant, the lack of systematic methods for
linking \textit{k}-mers to sequence variation hindered our ability to do so.

@@

% Seed coat color - G locus
We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all four
approaches (Figure \ref{seed_coat_color_greenyellow_manhattan}).
\cite{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant at that locus.  Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
Platypus analysis as corresponding to that causal variant (Figures
\ref{seed-coat-color-main-figure},
\ref{seed_coat_color_greenyellow_G_signal},
\ref{seed_coat_color_greenyellow_G_gene}). Unsurprisingly, none of the
SV-based GWAS analyses identified the causal variant given its nature.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the
\textit{k}-mer and Platypus analyses performed best since the region defined by the most highly associated markers
(top 1\%) included the causal gene (Figures
\ref{stem_termination_all_manhattan}, \ref{stem_termination_all_Dt1_signal}).
We did identify one of the SNPs suggested as putative causal variants for this
locus by \cite{tian2010} and \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Figures
\ref{stem_termination_all_Dt1_gene}, \ref{stem_termination_all_Dt1_kmers}).
Our results are consistent with those of \cite{liu2010} and \cite{tian2010} in
suggesting that there is more than one causal variant at that locus. Indeed, a
few cultivars bearing the allele associated with \emph{dt1} showed
indeterminate or semi-determinate phenotypes and some cultivars showed a
determinate phenotype without bearing the allele associated with \emph{dt1}
(Figure \ref{stem_termination_all_Dt1_kmers}). The existence of several causal
variants might explain why our analyses did not identify a causal variant more
clearly.

@@

The \textit{T} locus mentioned above for pubescence color is also a determining
locus for hilum color, and we detected signals overlapping the corresponding gene
using all approaches but vg (Figures \ref{hilum_color_all_manhattan},
\ref{hilum_color_all_T_signal}). The regions defined by the top 5\% (Paragraph)
and top 1\% (Platypus and \textit{k}-mers) variants/\textit{k}-mers included
the Glyma.06g202300 gene associated with this locus. Similarly to what we
observed for the analysis of the \textit{T} locus based on pubescence color, we
did identify the documented causal variant at this locus using the Platypus and
\textit{k}-mer analyses, although these variants were not the most
significantly associated (Figures \ref{hilum_color_all_T_gene},
\ref{hilum_color_all_T_kmers}).

@@

Similarly, the \textit{I} locus discussed above for its effect on seed coat
color also controls hilum color and we detected signals overlapping the
associated inverted duplication using all approaches but vg (Figures
\ref{hilum_color_all_manhattan}, \ref{hilum_color_all_I_signal}). However,
as was the case for the effect of \textit{I} on seed coat color, we were
not able to link any of the variants identified to the causal variation given
its highly complex nature.

@@

We conducted a GWAS analysis contrasting accessions with black and brown hila
in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected signals overlapping this locus using all approaches,
although these signals overlapped the Glyma.09g235100 gene associated with the
\textit{R} locus only for Platypus and \textit{k}-mers (Figures
\ref{hilum_color_blackbrown_manhattan}, \ref{hilum_color_blackbrown_R_signal}).
We observed two causal variants previously documented by \cite{gillman2011},
but these were not among the most significantly associated variants or
\textit{k}-mers (Figures \ref{hilum_color_blackbrown_R_gene},
\ref{hilum_color_blackbrown_R_kmers}).  This failure to identify causal
variants more directly is probably due to the existence of more than one causal
variant at this locus. Interestingly, the most significantly associated
\textit{k}-mers mapped just upstream of a gene (Glyma.09g234900) identified as
a homolog of Glyma.09g235100 \citep{gillman2011}. This result highlights how
difficult it may be to link significant \textit{k}-mers to causal variation in
some cases.

@@

We conducted a GWAS analysis contrasting accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. We
detected signals at the \textit{Ps} locus using all four approaches (Figures
\ref{pubescence-density-main-figure}a-b, \ref{pubescence_density_manhattan}). The causal
variant at this recently cloned locus is a copy number variant overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \cite{liu2020ps}. We found that
the most significantly associated \textit{k}-mer was located on the boundary of this CNV
and that reads containing this \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}b,d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

@@

We conducted a GWAS analysis contrasting dull and shiny seed coat phenotypes
for the seed coat luster trait. Seed coat luster was shown by \cite{gijzen1999}
to be largely caused by the deposition of a hydrophobic protein (HPS)
at the seed surface. Copy number variation of this gene at the \textit{B}
locus was later shown to explain variation in luster, although sequence
variation in the HPS sequence may also play a role \citep{gijzen2006}.
We found signals overlapping the causal CNV using all approaches, and the
most significantly associated \textit{k}-mer was notably located
within the CNV region (Figures \ref{seed_coat_luster_dullshiny_manhattan},
\ref{seed_coat_luster_dullshiny_B_signal}). However, we were not able
to directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

We did not detect any signals associated with maturity using our dataset
(Figure \ref{maturity_group_manhattan}). The use of maturity group as a proxy
for time to maturity instead of the number of days, combined with the limited
phenotypic variation at this trait in this collection of accessions (Table
\ref{MATURITY.GROUP_gwas_table}), likely explains our failure to detect
signals.

@@

\subsection*{Analysis of loci with unknown causal genes}
\label{sv-gwas-results}

@@

In addition to analyzing loci for which the genes are already known, we were
able to suggest candidate genes at loci for which the underlying genes are not
known yet. For example, both the \textit{L1} and the \textit{L2} loci for pod
color have yet to be cloned. While \cite{he2015} proposed the gene
Glyma.19g101700 as likely corresponding to \textit{L1} based on fine mapping
and expression analysis, our results as well as those reported by
\cite{bandillo2017} do not support this gene as a candidate due to its location
$\sim$3 Mb away from GWAS signals. Our analyses based on
\textit{k}-mers and Platypus found signals overlapping 14 and 6
genes, respectively (Figures \ref{pod_color_all_manhattan},
\ref{pod_color_all_L1_signal}). Of these, a MATE transporter (Glyma.19g120300)
appears as a prime candidate as the most significant \emph{k}-mer was observed
within its sequence and it is highly expressed in pods according to data hosted
on SoyBase \citep{grant2010, severin2010}. Moreover, the role of some MATE
proteins in transporting flavonoids has been demonstrated \citep{chen2015}.

The signals found by Platypus and \emph{k}-mer analyses at the \textit{L2}
locus overlap 24 and 50 genes, respectively (Figures
\ref{pod_color_all_manhattan}, \ref{pod_color_all_L2_signal}). Despite this
number of genes being somewhat high, we noticed that the closest gene to the
most significant \emph{k}-mer (Glyma.03g005800) is also a MATE transporter,
similarly to our top candidate for \emph{L1}. This gene may therefore represent
an interesting candidate for this locus.

@@

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, also have not been cloned yet, although \cite{gilbert2017} did
suggest Glyma.12g213900 as a candidate for \textit{Pa1} based on data published
by \cite{bandillo2017}. This gene encodes a Myb transcription factor annotated
as playing a role in trichome branching and is located near the most
significant signals found by the Platypus, Paragraph and \emph{k}-mer GWAS
analyses (Figures \ref{pubescence-form-main-figure}a-b,
\ref{pubescence_form_all_manhattan}, \ref{pubescence_form_all_Pa1_signal}).
GWAS analyses based on Platypus and \emph{k}-mers identified two highly
significant non-synonymous SNPs associated with appressed pubescence in this
gene (Figures \ref{pubescence-form-main-figure}c-d, \ref{pubescence_form_all_Pa1_gene},
\ref{pubescence_form_all_Pa1_kmers}). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Figures \ref{pubescence_form_all_manhattan},
\ref{pubescence_form_all_Pa2_signal}), none of which appeared obviously linked
to pubescence form.

@@

We used corrected dry weight (CDW) as described by \cite{deronne2022} as a
measure of horizontal resistance to \emph{P. sojae}. We found a strong signal
overlapping the genomic region reported by these authors on Gm15 using all four GWAS
approaches (Figure \ref{corrected_dry_weight_manhattan}). \cite{deronne2022}
proposed Glyma.15g217100 as a candidate gene for explaining phenotypic
variation at that locus based on a combination of functional annotation and
contrasting gene expression in resistant and susceptible lines four days after
inoculation. Although the signals found by our analyses did overlap with
Glyma.15g217100, we found that this gene was located at one end of the signal
and was found in a region with markers or \emph{k}-mers showing a weaker
association (Figure \ref{corrected_dry_weight_cdwGm15_signal}).  Although the
large size of the signal means that the gene associated with the phenotype could
be located almost anywhere in this region, we identified a particularly
interesting region defined by the top 1\% associations found from the Platypus
analysis. This region overlaps only three annotated genes and also contains the
most significant \emph{k}-mer (Figure
\ref{corrected_dry_weight_cdwGm15_signal}), which was located 30 kb away from a
gene (Glyma.15g217700) annotated as a calcium ion binding protein with an
EF-hand domain. Given the important role played by calcium signaling in defense
responses \citep{zhang2014}, we suggest Glyma.15g217700 as an interesting
candidate gene to explain resistance to \emph{P. sojae} in this population.

@@

In a few cases, we were also able to document possibly novel loci from the
results of \textit{k}-mer GWAS. In a GWAS analysis contrasting accessions
with indeterminate and semi-determinate stem termination types, we observed
previously undocumented signals on chromosomes Gm11, Gm16 and Gm18 in
addition to the known \textit{Dt1} locus (Figure \ref{stem-termination-main-figure}a).
Analysis of the pairwise LD between significant \textit{k}-mers found no
LD between these signals, suggesting that they may represent \textit{bona fide}
loci associated with stem termination type
(Figure \ref{stem-termination-main-figure}b). Similarly, a \textit{k}-mer GWAS
analysis contrasting accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15 that showed no LD with \textit{k}-mers
at the \textit{L1} locus and may therefore represent a different locus (Figures
\ref{pod_color_blbr_manhattan}, \ref{pod_color_blbr_ld}).

>>>>

We analyzed a total of 13 traits using three different GWAS approaches, one
based on SNP and indel genotypes (Platypus), one based on SV genotypes (Paragraph),
and one based on the presence/absence of \textit{k}-mers.
Because two or three different GWAS analyses were conducted for some of the traits, a
total of 22 GWAS analyses were performed using each of the four approaches.
Given the large volume of data that this represents, we will focus here on the
most noteworthy results. Readers interested in detailed results for loci not
discussed here are referred to Supplemental data. In particular, table \ref{signals_table}
provides a summary of the \textit{p}-values obtained by each approach at each
of the loci considered in this study.

@@

We first wanted to assess the performance of the various GWAS
approaches by analyzing the results at loci for which the genes have
already been cloned. These loci provide interesting test cases because
the expected results (causal genes or variants) are known. Table \ref{loci-table} shows
a summary of the results discussed below.

@@

\cite{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Figures
\ref{flower_color_manhattan}, \ref{flower_color_W1_signal}). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figures \ref{flower-color-main-figure}a-b and \ref{flower_color_W1_gene}).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure \ref{flower-color-main-figure}c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, variants located in the gene proved to be among the most closely
associated, but the causal variant could not be captured given the large size
of this variant.

We should note that when we first analyzed the results of the GWAS on flower
color, we observed that two accessions bearing the haplotype typically
associated with purple flowers were noted as having white flowers. Upon further
inspection, we found that the most likely cause of this discrepancy was a
mismatch between the genotype data and the identity of the accession for which
the phenotypic data was obtained. Indeed, when we compared SNP calls made
directly from the WGS data to those made using the SoySNP50K array, we observed
that these two accessions had low concordance rates of only 62\% and 70\%. To
ensure that such mismatches would not affect our GWAS analyses, we computed the
concordance between genotype calls made from WGS data and the SoySNP50K array
for all samples (see the Methods section) and
excluded 24 accessions for which the concordance rate was below 90\%.

@@

% Pubescence color - T locus
\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as
a putative causal variant at this locus. For all three GWAS approaches, we
found a signal overlapping this gene (Figure
\ref{pubescence_color_all_manhattan}). In
particular, the regions defined by the top 1\% most associated variants identified by the
Platypus and \textit{k}-mer approaches included Glyma.06g202300 (Figure
\ref{pubescence_color_all_T_signal}). While the most significant
\textit{k}-mers did occur within the sequence of this gene, they did not
correspond to the causal variant (Figure \ref{pubescence_color_all_T_gene}).
Still, the fourth and fifth most strongly associated \textit{k}-mers
corresponded to the causal variant (Figures \ref{pubescence_color_all_T_gene},
\ref{pubescence_color_all_T_kmers}). As for Platypus, the most significant
variant within the sequence of Glyma.06g202300 was the causal
variant, however it was not the most significantly associated variant overall
(Figures \ref{pubescence_color_all_T_signal},
\ref{pubescence_color_all_T_gene}). Paragraph did not detect the causal
variant at this locus given that this approach does not target small indels.

@@

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny and light
tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we
detected signals in this region using all four approaches, but only Platypus
and the \textit{k}-mers identified a signal that overlapped Glyma.03g258700
(Figures \ref{pubescence-color-main-figure}a-b,
\ref{pubescence_color_nogray_manhattan}, \ref{pubescence_color_nogray_Td_signal}).
While the most significant \textit{k}-mer identified the causal variant documented
at this locus \citep{yan2020}, it was the fourth most significantly associated variant
for the Platypus genotype calls (Figures
\ref{pubescence-color-main-figure}b-d, \ref{pubescence_color_nogray_Td_gene}).
Interestingly, some of the accessions bearing the haplotype associated with tawny pubescence
displayed light tawny pubescence, suggesting that there may be more than one causal
variant at this locus (Figure \ref{pubescence-color-main-figure}e).

@@

% Seed coat color - I locus
The causal variant at the \textit{I} locus for seed coat color is a complex SV
that takes the form of a 10.91-kb inverted duplication of three chalcone
synthase (CHS) genes \citep[CHS1, CHS3 and CHS4;][]{tuteja2008}. The presence of
this inverted duplication results in the silencing of all CHS genes in the seed
coat specifically and thus in the absence of seed coat pigmentation
\citep{tuteja2009}. While we did identify signals overlapping this causal variant
using all three GWAS approaches (Figures \ref{seed_coat_color_all_manhattan},
\ref{seed_coat_color_all_I_signal}), none of the approaches identified
the causal variant. This failure to detect the causal variant is likely due
to its complexity, which makes it difficult to detect and genotype it. While it
is likely that at least some of the significantly associated \textit{k}-mers
actually are derived from this causal variant, the lack of systematic methods for
linking \textit{k}-mers to sequence variation hindered our ability to do so.

@@

% Seed coat color - G locus
We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Figure \ref{seed_coat_color_greenyellow_manhattan}).
\cite{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
Platypus analysis as corresponding to that causal variant (Figures
\ref{seed-coat-color-main-figure},
\ref{seed_coat_color_greenyellow_G_signal},
\ref{seed_coat_color_greenyellow_G_gene}). Unsurprisingly, Paragraph
did not identify the causal variant given its type.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the
\textit{k}-mer and Platypus analyses performed best since the region defined by the most highly associated markers
(top 1\%) included the causal gene (Figures
\ref{stem_termination_all_manhattan}, \ref{stem_termination_all_Dt1_signal}).
We did identify one of the SNPs suggested as putative causal variants for this
locus by \cite{tian2010} and \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Figures
\ref{stem_termination_all_Dt1_gene}, \ref{stem_termination_all_Dt1_kmers}).
Our results are consistent with previous findings in
suggesting that there is more than one causal variant at the \textit{Dt1} locus \citep{liu2010, tian2010}. Indeed, a
few cultivars bearing the allele associated with \emph{dt1} showed
indeterminate or semi-determinate phenotypes and some cultivars showed a
determinate phenotype without bearing the allele associated with \emph{dt1}
(Figure \ref{stem_termination_all_Dt1_kmers}). The existence of several causal
variants might explain why our analyses did not identify a causal variant more
clearly.

@@

The \textit{T} locus mentioned above for pubescence color is also a determining
locus for hilum color, and we detected signals overlapping the corresponding gene
using all approaches but vg (Figures \ref{hilum_color_all_manhattan},
\ref{hilum_color_all_T_signal}). The regions defined by the top 5\% (Paragraph)
and top 1\% (Platypus and \textit{k}-mers) variants/\textit{k}-mers included
the Glyma.06g202300 gene associated with this locus. Similarly to what we
observed for the analysis of the \textit{T} locus based on pubescence color, we
did identify the documented causal variant at this locus using the Platypus and
\textit{k}-mer analyses, although these variants were not the most
significantly associated overall (Figures \ref{hilum_color_all_T_gene},
\ref{hilum_color_all_T_kmers}).

@@

Similarly, the \textit{I} locus discussed above for its effect on seed coat
color also controls hilum color and we detected signals overlapping the
associated inverted duplication using all approaches (Figures
\ref{hilum_color_all_manhattan}, \ref{hilum_color_all_I_signal}). However,
as was the case for the effect of \textit{I} on seed coat color, we were
not able to link any of the variants identified to the causal variation given
its highly complex nature.

@@

We conducted a GWAS analysis that only compared accessions with black and brown hila
in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches,
but signals overlapped the Glyma.09g235100 gene associated with the
\textit{R} locus only for Platypus and \textit{k}-mers (Figures
\ref{hilum_color_blackbrown_manhattan}, \ref{hilum_color_blackbrown_R_signal}).
We observed two causal variants previously documented by \cite{gillman2011},
but these were not among the most significantly associated variants or
\textit{k}-mers (Figures \ref{hilum_color_blackbrown_R_gene},
\ref{hilum_color_blackbrown_R_kmers}).  This failure to identify causal
variants more directly could be due to the existence of more than one causal
variant at this locus. Interestingly, the most significantly associated
\textit{k}-mers mapped just upstream of a gene (Glyma.09g234900) identified as
a homolog of Glyma.09g235100 \citep{gillman2011}.

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figures
\ref{pubescence-density-main-figure}a-b, \ref{pubescence_density_manhattan}). The causal
variant at this recently cloned locus is a copy number variant overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \cite{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}b,d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

The difference in size between the CNV reported by \cite{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 'N' nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any SV overlapping 'N' nucleotides in
the reference, we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, found that a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

Next, we conducted a GWAS analysis comparing dull and shiny seed coat phenotypes
for the seed coat luster trait. Seed coat luster has been shown by \cite{gijzen1999}
to be largely caused by the deposition of a hydrophobic protein (HPS)
at the seed surface. Copy number variation of this gene at the \textit{B}
locus was later shown to explain variation in luster, although sequence
variation in the HPS sequence may also play a role \citep{gijzen2006}.
We found signals overlapping the causal CNV using all approaches, and the
most significantly associated \textit{k}-mer was notably located
within the CNV region (Figures \ref{seed_coat_luster_dullshiny_manhattan},
\ref{seed_coat_luster_dullshiny_B_signal}). However, we were not able
to directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

We did not detect any signals associated with maturity using our dataset
(Figure \ref{maturity_group_manhattan}). The use of maturity group as a proxy
for time to maturity instead of the number of days, combined with the limited
phenotypic variation of this trait in this collection of accessions (Table
\ref{MATURITY.GROUP_gwas_table}), likely explains our failure to detect
any signals.

@@

\subsection*{Analysis of loci with unknown causal genes}
\label{sv-gwas-results}
In addition to analyzing loci for which the genes are already known, we were
able to suggest candidate genes at loci for which the underlying genes are not
known yet.

@@

Both the \textit{L1} and the \textit{L2} loci for pod
color have yet to be cloned. While \cite{he2015} proposed the gene
Glyma.19g101700 as likely corresponding to \textit{L1} based on fine mapping
and expression analysis, our results as well as those reported by
\cite{bandillo2017} do not support this gene as a candidate due to its location
$\sim$3 Mb away from GWAS signals. Our analyses based on
\textit{k}-mers and Platypus found signals overlapping 14 and 6
genes, respectively (Figures \ref{pod_color_all_manhattan},
\ref{pod_color_all_L1_signal}). Of these, a MATE transporter (Glyma.19g120300)
appears as a prime candidate as the most significant \emph{k}-mer was observed
within its sequence and it is highly expressed in pods according to data hosted
on SoyBase \citep{grant2010, severin2010}. Moreover, the role of some MATE
proteins in transporting flavonoids has been demonstrated \citep{chen2015}.

The signals found by Platypus and \emph{k}-mer analyses at the \textit{L2}
locus overlap 24 and 50 genes, respectively (Figures
\ref{pod_color_all_manhattan}, \ref{pod_color_all_L2_signal}). Despite this
somewhat high number, we noticed that the closest gene to the
most significant \emph{k}-mer (Glyma.03g005800) is also a MATE transporter,
similarly to our top candidate for \emph{L1}. This gene may therefore represent
an interesting candidate for this locus for future studies.

@@

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet, although \cite{gilbert2017} has suggested
Glyma.12g213900 as a candidate for \textit{Pa1} based on data published
by \cite{bandillo2017}. This gene encodes a Myb transcription factor annotated
as playing a role in trichome branching. Indeed, the most
significant signals found by all three approaches were located near this gene.
analyses (Figures \ref{pubescence-form-main-figure}a-b,
\ref{pubescence_form_all_manhattan}, \ref{pubescence_form_all_Pa1_signal}).
The analyses based on Platypus and \emph{k}-mers identified two highly
significant non-synonymous SNPs in Glyma.12g213900 that were associated with appressed pubescence
(Figures \ref{pubescence-form-main-figure}c-d, \ref{pubescence_form_all_Pa1_gene},
\ref{pubescence_form_all_Pa1_kmers}). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Figures \ref{pubescence_form_all_manhattan},
\ref{pubescence_form_all_Pa2_signal}), none of which appeared obviously linked
to pubescence form.

@@

We used corrected dry weight (CDW) as described by \cite{deronne2022} as a
measure of horizontal resistance to \emph{P. sojae}. We found a strong signal
overlapping the genomic region reported by these authors on Gm15 using all four GWAS
approaches (Figure \ref{corrected_dry_weight_manhattan}). \cite{deronne2022}
proposed Glyma.15g217100 as a candidate gene for explaining phenotypic
variation at that locus based on a combination of functional annotation and
contrasting gene expression in resistant and susceptible lines four days after
inoculation. Although the signals found by our analyses did overlap
Glyma.15g217100, we found that this gene was located at the distal end of the signal
in a region with markers or \emph{k}-mers showing a weaker
association (Figure \ref{corrected_dry_weight_cdwGm15_signal}). The
large size of the observed signal means that the gene associated with the phenotype could
be located almost anywhere in this region, however, we identified a particularly
interesting region defined by the top 1\% associations in the Platypus
analysis. This region overlaps only three annotated genes and also contains the
most significant \emph{k}-mer (Figure
\ref{corrected_dry_weight_cdwGm15_signal}). One of the genes in this region is
a calcium ion binding protein with an EF-hand domain (Glyma.15g217700).
Given the important role played by calcium signaling in defense
responses \citep{zhang2014}, we suggest Glyma.15g217700 as an interesting
candidate gene for explaining resistance to \emph{P. sojae} in this population.

@@

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.
When comparing accessions
with indeterminate and semi-determinate stem termination types, we observed
previously undocumented signals on chromosomes Gm11, Gm16 and Gm18 in
addition to the known \textit{Dt1} locus (Figure \ref{stem-termination-main-figure}a).
Analysis of the pairwise LD between significant \textit{k}-mers found no
LD between these signals, suggesting that they may represent \textit{bona fide}
loci associated with stem termination type
(Figure \ref{stem-termination-main-figure}b). Similarly, a \textit{k}-mer GWAS
analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15. \textit{k}-mers at this new locus showed no LD with \textit{k}-mers
at the known \textit{L1} locus and may therefore represent a distinct locus (Figures
\ref{pod_color_blbr_manhattan}, \ref{pod_color_blbr_ld}).
